
# **Parsing of storm wave and tidal data, and conversion to a single time-series**
---------------------------------------------------------------------------------

*Gareth Davies, Geoscience Australia 2017*

# Introduction
------------------

This document follows on from [preprocess_data.md](preprocess_data.md) in describing
our analysis of storm waves at Old Bar. It illustrates the extraction
of storm wave events from the "Old Bar" time-series created in the earlier script.

It is essential that the earlier script has already been run, and produced an RDS file - so we check
that now with code.
```{r checkorder}
# Check that the pre-requisites exist
if(!file.exists('Rimages/Session_data_processing_clean.Rdata')){
    stop('It appears you have not yet run preprocess_data.Rmd. That must be run before continuing')
}
```

Supposing the above did not generate any errors, and you have R installed,
along with all the packages required to run this code, and a copy of the
*stormwavecluster* git repository, then you should be able to re-run the
analysis here by simply copy-pasting the code. Alternatively, it can be run
with the `knit` command in the *knitr* package: 
```{r useknitr, eval=FALSE}
library(knitr)
knit('extract_storm_events.Rmd')
```

The basic approach followed here is to:
*Step 1: Extract storm events from the "Old Bar" time-series created earlier
*Step 2: Parse associated climate data, to support subsequent analysis.

Later we will develop the statistical analysis of the storm events.

# **Step 1: Extract storm events from the "Old Bar" time-series**
-----------------------------------------------------------------



