
# **Modelling the distributions of storm event statistics**
--------------------------------------------------------------------------

*Gareth Davies, Geoscience Australia 2017*

# Introduction
------------------

This document follows on from
[statistical_model_storm_timings.md](statistical_model_storm_timings.md)
in describing our statistical analysis of storm waves at Old Bar. 

It illustrates the process of fitting probability distributions to the storm event summary statistics,
which are conditional on the time of year and ENSO.

It is essential that the code
[statistical_model_storm_timings.md](statistical_model_storm_timings.md) has
alread been run, and produced an Rdata file
*'Rimages/session_storm_timings_FALSE_0.Rdata'*. **To make sure, the code below
throws an error if the latter file does not exist.**
```{r checkorder}
# Check that the pre-requisites exist
if(!file.exists('Rimages/session_storm_timings_FALSE_0.Rdata')){
    stop('It appears you have not yet run the code in statistical_model_storm_timings.md. It must be run before continuing')
}
```
You might wonder why the filename contains the term `FALSE_0`. Here, `FALSE`
describes where or not we perturbed the storm summary statistics before running
the fitting code. In the `FALSE` case we didn't - we're using the raw data.
However, it can be desirable to re-run the fitting code on perturbed data, to
check the impact of data discretization on the model fit. One would usually do
many such runs, and so we include a number (`_0` in this case) to distinguish
them.

Supposing the above did not generate any errors, and you have R installed,
along with all the packages required to run this code, and a copy of the
*stormwavecluster* git repository, then you should be able to re-run the
analysis here by simply copy-pasting the code. Alternatively, it can be run
with the `knit` command in the *knitr* package: 
```{r useknitr, eval=FALSE}
library(knitr)
knit('statistical_model_univariate_distributions.Rmd')
```

The basic approach followed here is to:
* **Step 1: Blah**

Later we will model the remaining joint dependence between these variables, and
simulate synthetic storm sequences. 

# **Step 1: Blah **
----------------------------------------------------------------------

