# Sensitivity of the fitted univariate distributions to random perturbations of the data
---------------------------------------------------------------------------------------

The code below investigates how fitted model parameters and AIC-based choice of
copula can change due to data perturbation.

**Read the model results, and do some basic checks**
```{r checkCode}

source('get_Rimage_data_univariate_distributions.R', local=TRUE)

# Read the summary statistics -- saved earlier for speed
store_var_list = readRDS('univariate_runs_summary_statistics.RDS')

# Read the original fit (based on un-perturbed data)
original_var_list = get_Rimage_data(
    '../statistical_model_fit/Rimages/session_univariate_distributions_FALSE_0.Rdata')

# Check that all the perturbed data sessions do jittering
stopifnot(all(sapply(store_var_list, f<-function(x) x$break_ties_with_jitter)))
# Check the original fit does not do jittering
stopifnot(original_var_list$break_ties_with_jitter == FALSE)

# Check that the event_statistics is unique in every session [i.e. the perturbed
# sessions really do randomly perturb event_statistics. We perturbed hsig, duration,
# tp1, and dir [tideResid was already unique].
#
# To do the check, compute the column sums of all event statistics. They should
# all be unique
#
max_es_vals = sapply(store_var_list, 
    f<-function(x) colSums(x$event_statistics[,1:4], na.rm=TRUE))
stopifnot(length(unique(max_es_vals)) == length(max_es_vals))


#
# Function to summarise parameters from perturbed runs
#
perturbed_summary<-function(variable_name){

    variable_vals = sapply(store_var_list,
        f<-function(x){
            num = x[[variable_name]] 
            denom = original_var_list[[variable_name]]
            # Handle errors gracefully
            if( (length(num) != length(denom)) || any(is.na(num)) || 
                (class(num) != class(denom))){
                num = denom*NA
            }
            return(num)
        }
    )

    variable_vals = t(variable_vals)
    print(summary(variable_vals))
    return(invisible())

}

#
# Function to summarise "relative errors" in perturbed runs
# i.e. (perturbed - original)/abs(original)
#
relative_error_summary<-function(variable_name){
    variable_differences = sapply(store_var_list,
        f<-function(x){
            num = x[[variable_name]] - original_var_list[[variable_name]]
            denom = original_var_list[[variable_name]]
            # Handle errors gracefully
            if( (length(num) != length(denom)) || any(is.na(num)) || 
                (class(num) != class(denom))){
                num = denom*NA
            }
            return(num/abs(denom))
        }
    )

    variable_differences = t(variable_differences)
    print(summary(variable_differences))
    return(invisible())
}
```



**Hsig model checks**

```{r hsig, fig.width = 8, fig.height=8}
# Make a plot
plot_ylim = range(c(original_var_list$hsig_mixture_fit_bayes_lower_q,
    original_var_list$hsig_mixture_fit_bayes_upper_q))

xs = original_var_list$hsig_mixture_fit_bayes_rates
plot(xs, original_var_list$hsig_mixture_fit_ml, 
    t='l', ylim=plot_ylim, main='Hsig fit (all overplotted)', 
    xlim=c(max(xs), min(xs)),
    log='xy', xlab='Event rate (number per year)', ylab='m')
points(xs, original_var_list$hsig_mixture_fit_bayes_median_q, col='orange', t='l', lty='dotted')
points(xs, original_var_list$hsig_mixture_fit_bayes_upper_q, col='red', t='l', lty='dotted')
points(xs, original_var_list$hsig_mixture_fit_bayes_lower_q, col='red', t='l', lty='dotted')
for(i in 1:length(store_var_list)){
    xs = store_var_list[[i]]$hsig_mixture_fit_bayes_rates
    points(xs, store_var_list[[i]]$hsig_mixture_fit_ml, t='l')
    points(xs, store_var_list[[i]]$hsig_mixture_fit_bayes_median_q, col='orange', t='l')
    points(xs, store_var_list[[i]]$hsig_mixture_fit_bayes_upper_q, col='red', t='l')
    points(xs, store_var_list[[i]]$hsig_mixture_fit_bayes_lower_q, col='red', t='l')
}
points(xs, original_var_list$hsig_mixture_fit_ml, col='blue', t='p', pch=1, lty='dashed')
grid()
legend('topleft', 
    c('Maximum Likelihood', 'Bayesian Median', 'Bayesian 97.5%', 'Bayesian 2.5%', 
        'Unperturbed data ML'),
    col=c('black', 'orange', 'red', 'red', 'blue'), lwd=c(1,1,1,1,NA), 
    pch=c(NA, NA, NA, NA, 1))

# Check how the hsig_mixture_fit parameters vary due to jittering
#
# Original parameters
print(original_var_list$hsig_mixture_fit_par)
# Perturbed parameter summary
perturbed_summary('hsig_mixture_fit_par')
# Relative error in parameters summary
relative_error_summary('hsig_mixture_fit_par')
#

# Original 1/100 quantile
print(original_var_list$hsig_aep100_quantiles)
# Summary of perturbed 1/100 quantile
perturbed_summary('hsig_aep100_quantiles')
# Summary of relative error
relative_error_summary('hsig_aep100_quantiles')

# Variability in optimal season phi [units of years -- typical value O(1 day)]
hsig_season_phi_err = unlist(sapply(store_var_list, 
    f<-function(x) x$hsig_season_phi%%1 - original_var_list$hsig_season_phi%%1))
print(summary(hsig_season_phi_err))

# Look at the automatically chosen copula family -- possibly more than one is identified
print(original_var_list$hsig_season_copula$familyname)
hsig_copula_type = sapply(store_var_list, f<-function(x) as.character(x$hsig_season_copula$familyname))
print(table(hsig_copula_type))
```

**Duration model checks**

```{r durationplot, fig.height=8, fig.width=8}
# Make a plot
plot_ylim = range(c(original_var_list$duration_mixture_fit_bayes_lower_q,
    original_var_list$duration_mixture_fit_bayes_upper_q))

xs = original_var_list$duration_mixture_fit_bayes_rates
plot(xs, original_var_list$duration_mixture_fit_ml, 
    t='l', ylim=plot_ylim, main='Duration fit (all overplotted)', xlim=c(max(xs), min(xs)),
    log='xy', xlab='Event rate (number per year)', ylab='m')
points(xs, original_var_list$duration_mixture_fit_bayes_median_q, col='orange', t='l', lty='dotted')
points(xs, original_var_list$duration_mixture_fit_bayes_upper_q, col='red', t='l', lty='dotted')
points(xs, original_var_list$duration_mixture_fit_bayes_lower_q, col='red', t='l', lty='dotted')
for(i in 1:length(store_var_list)){
    xs = store_var_list[[i]]$duration_mixture_fit_bayes_rates
    points(xs, store_var_list[[i]]$duration_mixture_fit_ml, t='l')
    points(xs, store_var_list[[i]]$duration_mixture_fit_bayes_median_q, col='orange', t='l')
    points(xs, store_var_list[[i]]$duration_mixture_fit_bayes_upper_q, col='red', t='l')
    points(xs, store_var_list[[i]]$duration_mixture_fit_bayes_lower_q, col='red', t='l')
}
points(xs, original_var_list$duration_mixture_fit_ml, col='blue', pch=1)
grid()
legend('topleft', 
    c('Maximum Likelihood', 'Bayesian Median', 'Bayesian 97.5%', 'Bayesian 2.5%', 'Unperturbed data ML'),
    col=c('black', 'orange', 'red', 'red', 'blue'), lwd=c(1,1,1,1, NA), pch=c(NA, NA, NA, NA, 1))

# Check how the duration_mixture_fit parameters vary due to jittering
#
print(original_var_list$duration_mixture_fit_par)
#
# Perturbed parameters
perturbed_summary('duration_mixture_fit_par')
relative_error_summary('duration_mixture_fit_par')
#
# 1/100 AEP
print(original_var_list$duration_mixture_fit_par)
perturbed_summary('duration_aep100_quantiles')
relative_error_summary('duration_aep100_quantiles')

# Errors in optimal season phi [units of years]
duration_season_phi_err = unlist(sapply(store_var_list, 
    f<-function(x) x$duration_season_phi%%1 - original_var_list$duration_season_phi%%1))
print(summary(duration_season_phi_err))

# Look at the automatically chosen copula family. Possibly a range of families are selected?
print(original_var_list$duration_season_copula$familyname)
duration_copula_type = unlist(sapply(store_var_list, f<-function(x) x$duration_season_copula$familyname))
print(table(duration_copula_type))
```

**Tidal residual model checks**


```{r tidal_residualA, fig.width=8, fig.height=8}
# Make a plot
plot_ylim = range(c(original_var_list$tideResid_mixture_fit_bayes_lower_q,
    original_var_list$tideResid_mixture_fit_bayes_upper_q))

xs = original_var_list$tideResid_mixture_fit_bayes_rates
plot(xs, original_var_list$tideResid_mixture_fit_ml, 
    t='l', ylim=plot_ylim, main='Tidal residual fit (all overplotted)', xlim=c(max(xs), min(xs)),
    log='xy', xlab='Event rate (number per year)', ylab='m')
points(xs, original_var_list$tideResid_mixture_fit_bayes_median_q, col='orange', t='l', lty='dotted')
points(xs, original_var_list$tideResid_mixture_fit_bayes_upper_q, col='red', t='l', lty='dotted')
points(xs, original_var_list$tideResid_mixture_fit_bayes_lower_q, col='red', t='l', lty='dotted')
for(i in 1:length(store_var_list)){
    xs = store_var_list[[i]]$tideResid_mixture_fit_bayes_rates
    points(xs, store_var_list[[i]]$tideResid_mixture_fit_ml, t='l')
    points(xs, store_var_list[[i]]$tideResid_mixture_fit_bayes_median_q, col='orange', t='l')
    points(xs, store_var_list[[i]]$tideResid_mixture_fit_bayes_upper_q, col='red', t='l')
    points(xs, store_var_list[[i]]$tideResid_mixture_fit_bayes_lower_q, col='red', t='l')
}
points(xs, original_var_list$tideResid_mixture_fit_ml, col='blue', t='p', pch=1)
legend('topleft', 
    c('Maximum Likelihood', 'Bayesian Median', 'Bayesian 97.5%', 
        'Bayesian 2.5%', 'Unperturbed data ML'),
    col=c('black', 'orange', 'red', 'red', 'blue'), lwd=c(1,1,1,1,NA), 
    pch=c(NA, NA, NA, NA, 1))

#
# Check how the tideResid_mixture_fit parameters vary due to jittering
#
print(original_var_list$tideResid_mixture_fit_par)
perturbed_summary('tideResid_mixture_fit_par')
relative_error_summary('tideResid_mixture_fit_par')
#
# 1/100 AEP 
print(original_var_list$tideResid_aep100_quantiles)
perturbed_summary('tideResid_aep100_quantiles')
relative_error_summary('tideResid_aep100_quantiles')

# Changes in seasonal phase
tideResid_season_phi_err = sapply(store_var_list, 
    f<-function(x) x$tideResid_season_phi%%1 - original_var_list$tideResid_season_phi%%1)
print(summary(tideResid_season_phi_err))

# Look at the automatically chosen copula family. 
print(original_var_list$tideResid_season_copula$familyname)
tideResid_copula_type = sapply(store_var_list, f<-function(x) x$tideResid_season_copula$familyname)
print(table(tideResid_copula_type))
```


**Steepness model checks**

Steepness is modelled non-parametrically.  
```{r steepnessModel, fig.width=8, fig.height=8}
#
# We use a non-parametric distribution, so here just check the copula and
# seasonal phase
#
plot(seq(0,1,len=200), original_var_list$steepness_quantiles_seq, t='l',
    main='Steepness', xlab='Non-exceedance probability for one event', ylab='Quantile')
for(i in 1:length(store_var_list)){
    points(seq(0,1,len=200), store_var_list[[i]]$steepness_quantiles_seq, t='l')
}
points(seq(0,1,len=200), original_var_list$steepness_quantiles_seq, pch=1, col='red')

legend('topleft', c('Non-parametric model', 'Unperturbed data model'), col=c('black', 'red'), 
    lwd=c(1,NA), pch=c(NA,1))

# Variability in seasonal phase
steepness_season_phi_err = sapply(store_var_list, 
    f<-function(x) (x$steepness_season_phi%%1 - original_var_list$steepness_season_phi%%1))
print(summary(steepness_season_phi_err))

# Variability in copula type
print(original_var_list$steepness_season_copula$familyname)
steepness_copula_type = sapply(store_var_list, f<-function(x) x$steepness_season_copula$familyname)
print(table(steepness_copula_type))
```

**Direction model checks**

Direction is modelled non-parametrically.  
```{r directionModel, fig.width=8, fig.height=8}
#
# We use a non-parametric distribution, so here just check the copula (soiA +
# seasonal) and seasonal phase
#
plot(seq(0,1,len=200), original_var_list$dir_quantiles_seq, t='l',
    main='Direction', xlab='Non-exceedance probability for one event', ylab='Quantile')
for(i in 1:length(store_var_list)){
    points(seq(0,1,len=200), store_var_list[[i]]$dir_quantiles_seq, t='l')
}
points(seq(0,1,len=200), original_var_list$dir_quantiles_seq, pch=1, col='red')
legend('topleft', c('Non-parametric model', 'Unperturbed data model'), col=c('black', 'red'), 
    lwd=c(1,NA), pch=c(NA,1))

# Variability in seasonal phase
dir_season_phi_err = sapply(store_var_list, 
    f<-function(x) (x$dir_season_phi%%1 - original_var_list$dir_season_phi%%1))
print(summary(dir_season_phi_err))

# Variability in direction/season copula family
print(original_var_list$dir_season_copula$familyname)
dir_copula_type = sapply(store_var_list, f<-function(x) x$dir_season_copula$familyname)
print(table(dir_copula_type))

# Variability in direction_soiA copula family
print(original_var_list$dir_soiA_copula$familyname)
dir_copula_type = sapply(store_var_list, f<-function(x) x$dir_soiA_copula$familyname)
print(table(dir_copula_type))

```
